--- 
title: "Introduction to R Shiny"
author: "Nitika K., Adele B., Paul F. H."
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
github-repo: nkandhari/R-shiny-MBP
description: "This is the course mateial for the workshop- `Introduction R Shiny` offered by [Monash Bioinformatics Platform](https://www.monash.edu/researchinfrastructure/bioinformatics)."
---

This workshop material is prepared by the [Monash Bioinformatics Platform](https://www.monash.edu/researchinfrastructure/bioinformatics) on the Shiny library developed by RStudio.

```{r}
library(bookdown)
```

# Prerequisites
Before the tutorial please install R and RStudio,

[Download R](https://cran.rstudio.com/)

[Download RStudio](https://www.rstudio.com/products/rstudio/download/)

Then install packages from CRAN and Bioconductor with the following R code:

```{r packages, eval=FALSE}
install.packages("shiny")
install.packages("DT")

source("https://bioconductor.org/biocLite.R")
biocLite(c(
    "BSgenome.Scerevisiae.UCSC.sacCer3",
    "TxDb.Scerevisiae.UCSC.sacCer3.sgdGene",
    "GenomicRanges",
    "rtracklayer",
    "Gviz"))
```

Make sure you are running R 3.3 and Bioconductor 3.3, or higher.

```{r bioc, eval=FALSE}
# To upgrade an old Bioconductor installation
source("https://bioconductor.org/biocLite.R")
biocLite("BiocUpgrade")
```

## Download files
* [tutorial.R](https://monashbioinformaticsplatform.github.io/2016-11-03-r-shiny/tutorial.R)

Download these files. tutorial.R contains all the code in the “tutorial” link below. During the tutorial, we will be stepping through tutorial.R in RStudio.

## Overview
* Introduction to Shiny

* Shiny structure

* Setup a basic Shiny app

* Tutorial: Let's make an app!

## Source code


```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), 'packages.bib')
```

